---
title: "R Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

############################################################
## Define packages to be loaded, and if needed, installed ##
##                                                        ##    
## by robert lohne (rlohne@gmail.com / @robertlohne)      ##
############################################################

# 1 Definitions
# 1.1 Package names 
# Define the names of the packages to be used, these are stored in the packages variable, and loaded/installed as needed
packages <- c("gapminder", "dplyr", "ggplot2", "gridExtra", "tidymodels", "ISLR", "mice", "visdat", "naniar", "parallel")

# 1.1 Packages description
# 

# 2. Install packages not yet installed
# Installed packages are stored in the installed packages variable
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# 3. Packages loading
invisible(lapply(packages, library, character.only = TRUE))
```

# 1. Introduction to logistic regression

Where linear regression is commonly used to predict continuous outcome variables, it can also be used for binary classifications. However, because it cannot easily predict qualitative outcome variables with more than two levels, another approach is needed. In this notebook, we'll look at logistic regression. There are indeed others as well, and we'll get to those in another notebook.

## 1.1 Binary outcomes

Let's first consider a classic classification task; a binary outcome. In this case, we will use the Default dataset, and predict whether or not customers default on their loan. We're going to work with what we have in the dataset, and also try to do some feature engineering.

In this example, we are dealing with a binomial, or binary logistic regression. This means that the outcome variable have only two possible types. A multinominal logistic regression is when the outcome variable can have three or more possible types, but they are not ordered (for instance if we tried to predict whether a car in the mtcars dataset was a SUV, a convertible or a sedan. The last type is the ordinal logistic regression, which is similar to the multinominal, but the levels are ordered. For instance if we tried to predict which Class a passenger traveled in on the Titanic.\
\
Mathematically, the formula for the probability of $Y$ being equal to 1 (Survived) can be expressed as:

$\log_b \frac{\rho}{1- \rho} = \beta_0 + B_1\chi_1 + B_2\chi_2$

where $log$ symbolizes the log-odds (logit), $_b$ is the base of the logarithm, and $\beta_1$, and $\beta_2$the parameters of the model, and $X$ the predictors.

## 1.2 Looking at our data

Let's look at our data. First, we take a look at the variables, their types and if any data are missing. To do this, I am going to use the vis_dat() function for the visdat-package. I like visdat because it has some very intuitive to use functions for visualizing a dataset - in particular when it comes to missing data. We start by loading the data and looking at it:

```{r visualize the Default dataset}
loans_df <- Default
vis_dat(loans_df)
```

As the figure above shows, we have two factor variables, default and student, and two numeric, income and
